Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-11-07 Writing/nw_main.Rnw
manuscript split into two: main and SI

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage[labelfont=bf, textfont=footnotesize]{caption} % \caption*{} will create legend (explanatory text)
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{authblk}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Aquifer depletion exacerbates agricultural drought losses in the US High Plains},
  pdfauthor={Taro Mieno; Timothy Foster; Shunkei Kakimoto; Nicholas Brozovic},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\author[1]{Taro Mieno}
\author[2]{Timothy Foster}
\author[3]{Shunkei Kakimoto}
\author[4]{Nicholas Brozovi\'{c}}
\affil[1]{Department of Agricultural Economics, University of Nebraska Lincoln}
\affil[2]{School of Mechanical, Aerospace and Civil Engineering, University of Manchester}
\affil[3]{Department of Applied Economics, University of Minnesota}
\affil[4]{Daugherty Water for Food Global Institute at the University of Nebraska}
\affil[4]{Department of Agricultural Economics, University of Nebraska Lincoln}

@@

Two major pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability \citep{foster2015analysis}. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs \citep{mieno2017price, bhattarai2021impact}. Second, depletion also reduces the thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation \citep{konikow2005groundwater, foster2014modeling, hrozencik2017heterogeneous}. If the reduction in well yields is substantial enough - for example as would occur when a large proportion of aquifer thickness is depleted \citep{hecox2002calculation, korus2020depletion} - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits \citep{foster2015well,rouhi2020downside}. The combined effect of declining well yields and increasing pumping costs could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production \citep{foster2014modeling, rad2020effects}. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change \citep{ukkola2020robust, chiang2021evidence, cook2020twenty}. Another potential pathway is through groundwater use regulation in response to low or declining aquifer thickness. However, it is unlikely that this effects is captured in our analysis. Please see Appendix \ref{pathway-reg}.

@@

Figure \ref{fig:yield-response} shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (panel A) and soybean (panel B). Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \ref{reg-conf} for estimated yield response curves with 95 percent confidence interval. The histogram of annual water deficit is presented at the top for each crop. In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases (see Figures \ref{fig:irrigated-yield-ind-corn} and \ref{fig:irrigated-yield-ind-soy} for the 95\% confidence intervals). This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors \citep{zhu2022untangling, li2020quantifying}.

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around \Sexpr{share_high} (\Sexpr{share_high * 100}\%) for an average county-level aquifer thickness of 100 m to \Sexpr{share_low} (\Sexpr{share_low * 100}\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, whether rainfed or irrigated is indicated by $ir$,  \(Y\) is crop yield, and \(D\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category-specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (see Appendix \ref{spline-basis} for further explanations on how spline-based regression works). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression to prevent our estimated response from becoming unrealistically oscillatory while maintaining a good fit based on Bayesian Information Criterion (BIC). To control for time-invariant heterogeneity across counties and yearly shocks, county-irrigation-status fixed effects $\alpha_{i, ir}$ and year fixed effects $\lambda_t$ are also included. County-irrigation-status fixed effects are used instead of county fixed effects because some counties have both rainfed and irrigated observations, and they need to be distinguished. Standard error estimation is clustered by county.

@@

where \(s_{i,t}\) is the share of irrigated production. The impact of average water deficit is captured by \(\sum_{k}^K \eta_{k}\phi_{k}(D_{i})\), where \(\phi_{k}(\cdot)\) is the \(k\)th spline basis function and \(\eta_{k}\) is its coefficient. The impact of aquifer thickness is captured in a similar way with \(\tau_{l}(\cdot)\) and \(\beta_{l}\) being the \(l\)th spline basis function and its coefficient, respectively. $K$ and $L$ are the numbers of basis functions for $D$ and $A$, respectively. $K = 4$ and $L = 4$ were chosen to balance preventing the estimated responses from being unrealistically oscillatory and ensuring a good fit as indicated by BIC. For this regression, counties with aquifer thickness levels greater than 180 m were excluded due to the sparsity of observations in that range because our flexible functional form approach cannot reliably estimate its impact. County-level soil characteristics including sand percentage, silt percentage, and water holding capacity, are represented by \(X_i\). State-year fixed effects and error term are represented by \(\theta_{s,t}\) and \(v_{i,t}\), respectively. Unlike for our regression models of per-area yields, we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects would effectively eliminate the cross-sectional variation in aquifer thickness. A formal discussion of the trade-offs associated with the choice of fixed effects is provided in Appendix \ref{county-fe}.

@@

Funding for the research in this manuscript was provided by the United States Department of Agriculture under contract numbers OCE 58-0111-20-007 and OCE 58-0111-21-007 and the Daugherty Water for Food Global Institute at the University of Nebraska. This research was also supported by Innovate UK award no. 10044695, as part of the UK Research and Innovation and European Commission funded project ``TRANSCEND: Transformational and robust adaptation to water scarcity and climate change under deep uncertainty.''

The funders had no role in study design, data collection and analysis, decision to publish or preparation of the manuscript.

@@

\caption*{Note: The solid lines represent per-area irrigated yield responses to water deficits, dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $\Sexpr{nobs_corn}$ and $\Sexpr{nobs_soy}$ for corn and soybean regressions, respectively.}
\label{fig:yield-response}
\end{figure}

@@

\caption*{Note: The solid red lines represent the difference in estimated per-area irrigated corn (top panels) and soybean (bottom panels) yields for the second and third (largest) aquifer thickness quantiles relative to the first (smallest) aquifer thickness quantile. Shaded area represents 95\% confidence interval.}
\label{fig:yield-dif}
\end{figure}

@@

\label{fig:ir-share}
\caption*{Note: The solid black line represents the estimated share of irrigated acres. Shaded area represents 95\% confidence interval. The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $\Sexpr{nobs_corn_share}$ and $\Sexpr{nobs_soy_share}$ for corn and soybean regressions, respectively.}
\end{figure}

@@

% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_total_impact}}
\caption*{Note: The solid lines represent average yield at different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix \ref{reg-conf} for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop.}
\label{fig:tot-impact}
\end{figure}

@@

\caption*{Note: The solid red lines represent difference in average yield of corn and soybean for aquifer thickness levels of 40 m, 70 m, and 100 m, relative to a baseline aquifer thickness of 10 m. Shaded areas denote 95\% confidence interval, which when greater than zero indicate average yields for the given aquifer thickness level are statistically significantly greater than for an aquifer thickness of 10 m at the 5\% level of significance.}
\label{fig:dif-tot-impact}
\end{figure}

@@

\caption*{Note: The blue solid, red solid, and black dotted lines represent irrigated yield, rainfed yield, and water deficit, respectively.}\label{fig:deficit-yield-hist}
\end{figure}

@@

\caption*{Note: The map shows county-average aquifer thickness in 2016 for each county included in our corn and soybean regression analyses. Counties highlighted in gray are rainfed only, and the red polygon denotes the boundary of the High Plains Aquifer system}\label{fig:sat-map}
\end{figure}

@@

%==============================================
%= Supplementary Information
%==============================================
\clearpage

\hypertarget{appendix-appendix}{%
\appendix}

\hypertarget{pathway-reg}{%
\section{Groundwater management policies as a potential causal pathway}\label{pathway-reg}}

Our analysis evaluates the links between aquifer thickness and drought resilience, proposing a causal link between declining aquifer thickness and irrigated production during droughts. In the first section of our manuscript, we propose that this connection is driven by impacts of aquifer thickness on well yields and, hence, farmers ability to use groundwater to fully buffer crops against water deficits during drought.

It is important to assess the extent to which this causal relationship may be affected or confounded by other factors. One key alternative hypothesis is that regulations limiting groundwater withdrawals could have been introduced as a response to declining aquifer thickness, with resulting reductions in irrigated areas and irrigated crop yields driven by regulatory rather than hydrologic (i.e. well yield) constraints. Below, we discuss the reasons that regulatory constraints are highly unlikely to be the cause of the relationships between aquifer thickness and drought impacts observed in our analysis for the HPA.

First, it is important the majority of groundwater withdrawal regulations that do exist in the HPA limit pumping to a specified volume over a multiyear period. It is also common for farmers to have been allowed to bank and carryover significant volumes of water deficits from historic periods when allocation levels were higher \citep{rimvsaite2023groundwater}, further increasing their regulatory quotas above defined caps. Together, these factors mean that farmers in parts of the HPA where groundwater withdrawals are regulated or restricted are still able to increase water withdrawals substantially in drought years when demand is higher. Prior research has shown that as a result farmers water use decisions in HPA are not constrained by such regulations even in extreme drought years such as 2012 \citep{foster2019assessing}.

Where more restrictive groundwater withdrawal limits have been introduced, for example in some Groundwater Management Districts in Kansas \citep{marston2022, rimvsaite2023groundwater}, these have universally been implemented in the last 5-10 years and therefore have no or only minimal overlap with our study period which extends up to only 2016. The earliest of these groundwater managements in Kansas interventions was the Sheridan 6 Local Enhanced Management Area (LEMA) in Northwest Kansas, which introduced restrictions to farmer groundwater pumping of 11 inches per year on average per 5-year period starting in 2013. While this overlaps with the final 4 years of our study period, the Sheridan 6 LEMA area covers only parts of two counties in Kansas and hence represents a very small proportion of our overall study sample. Additionally, the years 2013-2016 were all characterized by average or above average growing season rainfall (i.e. not drought years). Furthermore, evidence shows that the effect of restrictions introduced by the LEMA has primarily been a reduction in per-area water use, with irrigated crop yields largely unaffected and only small reductions in irrigated areas compared to pre-policy conditions \citep{drysdale2018adaptation, deines2021combining, whittemore2023108408}. These policies therefore would not be expected to confound our results on the impacts of aquifer thickness.

Finally, we also highlight that areas of the southern HPA, in particular in northern Texas, where the lowest aquifer thickness values are observed have few if any regulatory restrictions on groundwater extraction. This is due to laws that typically provide farmers with absolute rights to use groundwater under their land, coupled with low levels of metering. Research has shown that these conditions have historically limited or constrained potential for implementing extensive and binding pumping restrictions on irrigating farmers in these areas \citep{wheeler2016lessons, closas2018chronicle}, again indicating that regulations are not the binding driver of water use constraints in these areas. As discussed in section \ref{main} of our manuscript, there is extensive evidence pointing to reduced well yields in these areas of the HPA. This evidence ranges from anecdotal and theoretical insights to simulation-based studies. Collectively, they suggest that these restrictions significantly constrain farmers' ability to meet crop water demands during droughts.

\hypertarget{reg-conf}{%
\section{Uncertainties in estimated impacts of water deficit on irrigated and average yield}\label{reg-conf}}

\setcounter{figure}{0}
\renewcommand{\thefigure}{B.\arabic{figure}}

In the main article, the confidence intervals of the estimated yield responses to water deficit were not presented to enable easier comparison between response curves. This section provides visualization of yield response with the 95 \(\%\) confidence interval to convey the degree of uncertainty in our estimation. Figures \ref{fig:irrigated-yield-ind-corn} and \ref{fig:irrigated-yield-ind-soy} present the impact of water deficit on crop yield by aquifer thickness category with the 95\(\%\) confidence interval for irrigated corn and soybean, respectively. Figure \ref{fig:avg-yield-ind} presents the impact of water deficit on average crop yield (irrigated and rainfed combined) by aquifer thickness category for corn and soybean with the 95\(\%\) confidence interval.

%------------------------
%- Figure B.1
%------------------------

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area irrigated yields of corn}

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_yield_with_conf_corn}}

\caption*{Note: The solid black lines represent the estimated rainfed and irrigated per-area irrigated yields of corn in US High Plains at different values of water deficit and aquifer thickness. The gray shaded area represents the 95\% confidence interval.}
\label{fig:irrigated-yield-ind-corn}
\end{figure}

%------------------------
%- Figure B.2
%------------------------

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area irrigated yields of soybean}

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_yield_with_conf_soy}}

\caption*{Note: The solid black lines represent the estimated rainfed and irrigated per-area irrigated yields of soybean in US High Plains at different values of water deficit and aquifer thickness. The gray shaded area represents the 95\% confidence interval.}
\label{fig:irrigated-yield-ind-soy}
\end{figure}

%------------------------
%- Figure B.3
%------------------------

\caption{Average productivity of corn and soybean}

{\centering \includegraphics[width=6.5in,height=500px,]{../Results/Figures/g_avg_yield}}

\caption*{Note: The black solid lines represent average yield of corn (top panel) and soybean (bottom panel) in the US High Plains for different levels of water deficit and aquifer thickness. The gray shaded area represents the 95\% confidence interval.}
\label{fig:avg-yield-ind}
\end{figure}

\hypertarget{test-dif-share}{%
\section{Comparison of the impacts of different aquifer thickness levels on irrigated area share}\label{test-dif-share}}

\setcounter{figure}{0}
\renewcommand{\thefigure}{C.\arabic{figure}}

<<sat_ref, echo = FALSE>>=
sat_ref <-
  readRDS(here::here("Results/all_results.rds")) %>%
  .[1, sat_seq] %>%
  .[[1]] %>%
  .[length(.)]
@

Figure \ref{fig:test-sat-share} presents the difference in the estimated share of irrigated acres at different levels of aquifer thickness relative to the aquifer thickness level of \Sexpr{sat_ref} m, which represents a high level of aquifer thickness for our study area that exerts no constraints on water supply for irrigation. Gray shaded areas represented the 95\% confidence intervals for the estimated relationships. Where this range exceeds the red zero line, this indicates that the difference in aquifer thickness (relative to \Sexpr{sat_ref} m baseline) leads to a statistically significant (at 5\% level) reduction in irrigated production area share.

%------------------------
%- Figure C.1
%------------------------

\caption{Decline in the share of acres that are irrigated}

{\centering \includegraphics[width=6.5in,height=500px,]{../Results/Figures/g_share_dif}}

\caption*{Note: The solid black lines represent the share of acres that are irrigated as a function of aquifer thickness relative to a baseline aquifer thickness level of \Sexpr{sat_ref} m (black line). The gray shaded areas represent the 95\% confidence interval for this estimated relationship.}\label{fig:test-sat-share}
\end{figure}

\hypertarget{county-fe}{%
\section{Comparison of the choice of spatial fixed effects in irrigation area share regressions}\label{county-fe}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage[labelfont=bf]{caption} 
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
% \linenumbers
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{authblk}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\thispagestyle{empty}
\hypersetup{
  pdftitle={Aquifer depletion exacerbates agricultural drought losses in the US High Plains},
  pdfauthor={Taro Mieno; Timothy Foster; Shunkei Kakimoto; Nicholas Brozovic},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\author[1]{Taro Mieno}
\author[2]{Timothy Foster}
\author[3]{Shunkei Kakimoto}
\author[4]{Nicholas Brozovi\'{c}}
\affil[1]{Department of Agricultural Economics, University of Nebraska-Lincoln}
\affil[2]{School of Mechanical, Aerospace and Civil Engineering, University of Manchester}
\affil[3]{Department of Applied Economics, University of Minnesota}
\affil[4]{Daugherty Water for Food Global Institute at the University of Nebraska}
\affil[4]{Department of Agricultural Economics, University of Nebraska-Lincoln}

@@

Two major pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability \citep{foster2015analysis}. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs \citep{mieno2017price, bhattarai2021impact}. Second, depletion also reduces the thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation \citep{konikow2005groundwater, foster2014modeling, hrozencik2017heterogeneous}. If the reduction in well yields is substantial enough - for example as would occur when a large proportion of aquifer thickness is depleted \citep{hecox2002calculation, korus2020depletion} - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits \citep{foster2015well,rouhi2020downside}. The combined effect of declining well yields and increasing pumping costs could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production \citep{foster2014modeling, rad2020effects}. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change \citep{ukkola2020robust, chiang2021evidence, cook2020twenty}. Another potential pathway is through groundwater use regulation in response to low or declining aquifer thickness. However, it is unlikely that this effects is captured in our analysis. For more details on this issue, please refer to section A of the Supplementary Information.

@@

Figure \ref{fig:yield-response} shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (panel A) and soybean (panel B). Confidence intervals were not presented to avoid over-crowding the figures. Please see section B of the Supplementary Information for estimated yield response curves with 95 percent confidence interval. The histogram of annual water deficit is presented at the top for each crop. In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases (see Figures S.1 and S.2 for the 95\% confidence intervals). This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors \citep{zhu2022untangling, li2020quantifying}.

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around \Sexpr{share_high} (\Sexpr{share_high * 100}\%) for an average county-level aquifer thickness of 100 m to \Sexpr{share_low} (\Sexpr{share_low * 100}\%) for an average county-level aquifer thickness of 10 m (Please see section C of the Supplementary Information for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, whether rainfed or irrigated is indicated by $ir$,  \(Y\) is crop yield, and \(D\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category-specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (For further explanations on how spline-based regression functions, please see section D of the Supplementary Information.). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression to prevent our estimated response from becoming unrealistically oscillatory while maintaining a good fit based on Bayesian Information Criterion (BIC). To control for time-invariant heterogeneity across counties and yearly shocks, county-irrigation-status fixed effects $\alpha_{i, ir}$ and year fixed effects $\lambda_t$ are also included. County-irrigation-status fixed effects are used instead of county fixed effects because some counties have both rainfed and irrigated observations, and they need to be distinguished. Standard error estimation is clustered by county.

@@

where \(s_{i,t}\) is the share of irrigated production. The impact of average water deficit is captured by \(\sum_{k}^K \eta_{k}\phi_{k}(D_{i})\), where \(\phi_{k}(\cdot)\) is the \(k\)th spline basis function and \(\eta_{k}\) is its coefficient. The impact of aquifer thickness is captured in a similar way with \(\tau_{l}(\cdot)\) and \(\beta_{l}\) being the \(l\)th spline basis function and its coefficient, respectively. $K$ and $L$ are the numbers of basis functions for $D$ and $A$, respectively. $K = 4$ and $L = 4$ were chosen to balance preventing the estimated responses from being unrealistically oscillatory and ensuring a good fit as indicated by BIC. For this regression, counties with aquifer thickness levels greater than 180 m were excluded due to the sparsity of observations in that range because our flexible functional form approach cannot reliably estimate its impact. County-level soil characteristics including sand percentage, silt percentage, and water holding capacity, are represented by \(X_i\). State-year fixed effects and error term are represented by \(\theta_{s,t}\) and \(v_{i,t}\), respectively. Unlike for our regression models of per-area yields, we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects would effectively eliminate the cross-sectional variation in aquifer thickness. A formal discussion of the trade-offs associated with the choice of fixed effects is provided in section E of the Supplementary Information.

@@

Funding for the research in this article was provided by the United States Department of Agriculture under contract numbers OCE 58-0111-20-007 and OCE 58-0111-21-007 and the Daugherty Water for Food Global Institute at the University of Nebraska. This research was also supported by Innovate UK award no. 10044695, as part of the UK Research and Innovation and European Commission funded project ``TRANSCEND: Transformational and robust adaptation to water scarcity and climate change under deep uncertainty.''

The funders had no role in study design, data collection and analysis, decision to publish or preparation of the article.

@@

\caption*{\footnotesize{Note: The solid lines represent per-area irrigated yield responses to water deficits, dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $\Sexpr{nobs_corn}$ and $\Sexpr{nobs_soy}$ for corn and soybean regressions, respectively.}}
\label{fig:yield-response}
\end{figure}

@@

\caption*{\footnotesize{Note: The solid red lines represent the difference in estimated per-area irrigated corn (top panels) and soybean (bottom panels) yields for the second and third (largest) aquifer thickness quantiles relative to the first (smallest) aquifer thickness quantile. Shaded area represents 95\% confidence interval.}}
\label{fig:yield-dif}
\end{figure}

@@

\caption*{\footnotesize{Note: The solid black line represents the estimated share of irrigated acres. Shaded area represents 95\% confidence interval. The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $\Sexpr{nobs_corn_share}$ and $\Sexpr{nobs_soy_share}$ for corn and soybean regressions, respectively.}}
\label{fig:ir-share}
\end{figure}

@@

% {\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_total_impact}}
\caption*{\footnotesize{Note: The solid lines represent average yield at different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers' choices about irrigated area shares in response to different water deficit and aquifer conditions. Please see section B of the Supplementary Information for estimated average yield response curves with 95\% confidence intervals displayed, which are omitted here for visual clarity. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop.}}
\label{fig:tot-impact}
\end{figure}

@@

\caption*{\footnotesize{Note: The solid red lines represent difference in average yield of corn and soybean for aquifer thickness levels of 40 m, 70 m, and 100 m, relative to a baseline aquifer thickness of 10 m. Shaded areas denote 95\% confidence interval, which when greater than zero indicate average yields for the given aquifer thickness level are statistically significantly greater than for an aquifer thickness of 10 m at the 5\% level of significance.}}
\label{fig:dif-tot-impact}
\end{figure}

@@

\caption*{\footnotesize{Note: The blue solid, red solid, and black dotted lines represent irrigated yield, rainfed yield, and water deficit, respectively.}}
\label{fig:deficit-yield-hist}
\end{figure}

@@

\caption*{\footnotesize{Note: The map shows county-average aquifer thickness in 2016 for each county included in our corn and soybean regression analyses. Counties highlighted in gray are rainfed only, and the red polygon denotes the boundary of the High Plains Aquifer system.}}
\label{fig:sat-map}
\end{figure}

@@
